The Small World Network Effect in Software Project Teams \author{Kevin Peterson\\ \small \texttt{pete1968@umn.edu} }

### 2013-12-03 paper/main.tex
First proof-read run.

\begin{abstract}
Team cohesion and the dynamics of team forming are important parts of any project, with software projects being no exception. An interesting aspect of team building is the relationships formed between the team members. It follows then that visualizing software team members as a graph may be a natural way to explore these relationships. As team members move between projects, these graphs become more and more connected as team members collaborate and form relationships. We show that this connectivity, known as the ``small world effect," has a positive impact on team performance when the connectivity levels are moderate. Performance degrades, however, at both very high and very low levels of connectivity. This aligns with similar research findings of non-software teams.
\end{abstract}

@@

A ``small world'' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes\cite{watts1998collective}. ``Clustering'' in this context means how closely nodes in the graph are related to each other, or ``the friend of my friend is also my friend.'' Milgram, in is seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}, or in other words, the average ``path length'' between any two people.

Uzzi \cite{uzzi2005collaboration} expanded on this by exploring how these small world networks correlate to collaboration and creativity. In his study, the network of artists involved in the making of Broadway musicals from 1945 to 1989 were studied. Performance of the musicals were then measured based on financial success and critical acclaim, and compared that quantitative analysis to the clustering and path length of the graph.

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} data sets of free and open source projects. Two datasets were analyzed -- data from from the popular open source project site \textit{Freecode},\footnote{http://freecode.com/}, and another from open source software repository \textit{SourceForge}. The Freecode dataset contained data up to September, 2013, while the SourceForge dataset was slightly older with data gathered on June 2009. Both datasets were chosen because they met two basic criteria: (1) they contained an exhaustive list of projects and contributors to those projects, and (2) they provided some sort of quantitative measure of project ``popularity." This popularity metric will be further described in sectiions below. Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the contributors and the edges representing one or more shared project between contributors. For our purposes, ``contributors'' includes any person that has been identified as being associated with the project by the project hosting site (Freecode and SourceForge, for our purposes). It is important to note that contributor does not always imply code contributions -- as issue reporting and documentation, among other things, are valid contributions. Various properties of the graph were analyzed, followed by a qualitative section to help link the quantitative analysis to subject matter expert experiences.

\subsubsection{Graph Structure}
For each dataset, we first considered the dataset to be one large graph. From this large disconnected graph, each connected subgraph was analyzed. Metrics were then computed for each of these subgraphs. In order to correctly compute the metrics below, a minimum subgraph size was required. Specifically, many of the graph metrics of interest measured ``triangles'' of nodes and their connections. This implies a subgraph of {$>= 3$} nodes is needed. Subgraphs of {$< 3$} nodes were not considered for analysis.

@@

A bipartite graph projection is necessary to further analyze the graph. A bipartite projection involves taking the disjoint node sets $P$ and $C$, and representing the graph as relationships between only one of those sets of nodes. A projection onto the contributor nodes, or $C$, of figure \ref{fig:example_bipartite_graph} is show in figure \ref{fig:example_bipartite_projection_graph}. Here, contributors are connected directly, with edges occurring if a two contributors share one or more common projects. It is important to note that the projection could have been done in terms of the projects, not the contributors. This would have led to a graph with nodes of projects $P$, each being linked by sharing a common contributor. This approach is a common way of representing these bipartite graphs\cite{newman2001scientific}, and illustrates the fact that a bipartite projections are lossy compared to the data represented in the original graph\cite{zhou2007bipartite}.\\

@@

\subsubsection{Graph Analysis}
To support our hypotheses, there are several key metrics to be studied in regards to the aforementioned graph. Many of these metrics follow the work of Boccaletti \textit{et. al}\cite{boccaletti2006complex}, and other studies of similar graph types\cite{latora2001efficient,adamic1999small}.

@@

\noindent\\\textit{\textbf{Average Path Length ($L$)}}\\
Also, calculating the \textit{Path Length} ($L$) allows us to determine how far removed developers in the graph are from one another. The path length is calculated as the average distance between any two node pairs in the graph:

@@

Where {$v_i \in P$}, {$v_j \in P$}, and {$\lambda(v_i,v_j)$} denotes the shortest path between the two nodes. We only consider path length for connected subgraphs, avoiding some complexities of this calculation on disconnected graphs\cite{boccaletti2006complex}.

@@

\subsubsection{Qualitative Analysis}
So far we have explored, via data analysis, how closely project collaborators are connected to one another, and what correlations we can derive from this analysis. We have said little as to why these connections are made, how they are maintained, and what exactly a connection means at a personal level. Via qualitative analysis, we will attempt to enrich our data analysis by providing deeper context.\\
A connection between two people is much more than an edge between two nodes in a graph. It has been asserted that people tend to make connections with people with whom they share similarities\cite{mcpherson2001birds}. This idea of homophily, however, needs to be further expanded before we can make assertions about correlations to performance. It has been shown that demographic homophily is complex to measure, correlates little to performance outcomes, and in fact correlate negatively\cite{reagans2004make,lawrence1997perspective}. For our purposes, it is much more appropriate to examine how teams best leverage similarities of knowledge and information, as the flow of these assets can be a key driver to success\cite{nissen2002extended}.\\
In the same way we can explore the role of team familiarity. The two types of familiarity that we will explore are \textit{member} familiarity and \textit{task/project} familiarity\cite{harrison2003time}. These metrics, although not captured in data, could be important factors in \textit{how} teams form, and thus help us to understand \textit{why} teams perform the way that they do.\\

To further study these factors, we arranged interviews with two current professionals in the field. Both have had experience with project management and are certified Project Management Professionals (PMP)\textregistered. We proposed to them the following questions, in hopes of expanding further the questions posed above:

@@

Summary statistics for the graph analysis are captured in figure \ref{fig:summary_stats}.  Statistics analyzed included:\\
$Q$ - \textit{Small World Quotient}\\
$C^\Delta$ - \textit{Clustering Coefficient - Transitivity}\\
$\overline{C^\lambda}$ - \textit{Clustering Coefficient - Average}\\
$L$ - \textit{Average Path Length}\\
$N$ - \textit{Number of Nodes}\\
$E$ - \textit{Number of Edges}\\
As measured over the number of projects $P$, with the total number of contributors $C$. The number of connected subgraphs analyzed is shown as $P_c$. The connected subgraphs are the main unit of analysis, and each statistical operation is applied to the set of subgraphs $P_c$.

A mean value of $Q$ was observed to be $> 1$ for both data sets, which meets the criteria for a small-world network\cite{humphries2008network}. The median, however, is 1.0 in both cases, which does not match the small world network criteria. A possible explanation of this is the power law distribution of $Q$ observed in both data sets (figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}). This indicates that a large proportion of $Q$ values are around 1.\\

@@

Note that $P_c$ values are relatively small compared to $P$ and $C$ values. It would be intuitive to assume that {$N \times P_c \approx C$} -- in other words, if we split the entire graph into connected subgraphs, the number of subgraphs ($P_c$) multiplied by the average number of nodes ($N$) per subgraph should equal the total number of contributors ($C$). This, however, is not the case. In fact, $P_c$ is notably smaller than expected. Recall that during analysis, subgraphs of {$< 3$} nodes were not considered. We can assume then that a large number of subgraphs are very small, with only one or two nodes. Further analysis is needed to fully understand the implications of this.

To validate the hypothesis, we expect to find that intermediate values of $Q$ will be correlated with more popular projects. To explore this possibility, we first compare the popularity given $Q$ values as shown in figures \ref{fig:q_fc_graph} and \ref{fig:q_sf_graph}. The $Q$ values do display the U-shaped characteristic as compared to popularity, which aligns with the findings of Uzzi \textit{et. al}\cite{uzzi2005collaboration}.

@@

\subsection{Qualitative}
The research interviews provided a look into how some of these quantitative findings translate into actual industry practice. It also helped us to obtain a deeper understanding of what we had found in the data analysis, as we could begin to build some workplace context into our results. There were are few re-occurring themes from which we will base our analysis\cite{stmartin_interview,rana_interview}:

@@

Although our data analysis does not include any examination of a core/periphery, it is an interesting direction for future research. Notably, has been suggested that individual performance in a group is highest if that individual is not in the core or the periphery, but somewhere in between\cite{cattani2008core}. This is especially interesting as this aligns with our hypothesis that intermediate values of $Q$ are optimal.

\noindent\\\textit{\textbf{Teams can stagnate without new members.}}\\
Both interview participants indicated that new members were essential in bringing new ideas to the team. A highly static team (or, a team with a high $Q$ value), runs the risk of falling behind the technology trends. Occasional new members (or, intermediate values of $Q$) was perceived as a positive in terms of information flow

@@

\noindent\\\textit{\textbf{Priority, focus, and timeline are strong drivers.}}\\
Focus was specifically called out as an indicator of performance in one interview. Clear goals and a common objective were positive influencer, while continuously shifting focus was a negative influence. Research in this area suggests that creative teams have ``goals that are clear and compelling, but also open and challenging''\cite{isaksen2002climate}.

Timeline and priority discussed as strong contributors to team success. Team building could be thought of as function of the timeline, meaning that as the timeline of the project increased, more time could be devoted to team building activities.

@@

\section{Discussion and Future Research}
Social networks, interpersonal relationships, and software development are all complex topics. We have explored a narrow slice of these in context, and there are some important areas of future exploration. First, the contributor graph contains a large amount of subgraphs, and as a whole appears less clustered than some other observed social network graphs -- consistent with Madey's findings\cite{madey2002open} on SourceForge data. As noted in our analysis, a large number of these subgraphs have only one or two contributors. This implies that a large number of contributors work alone and on a single project. We do not, with the current analysis, know how (or if) this degrades the quality of our findings.\\

>>>>

\begin{abstract}
Team cohesion and the dynamics of team forming are important parts of any project, with software projects being no exception. An interesting aspect of team building is the relationships formed between the team members. Because of these relationships, visualizing software team members as a graph may be a natural way to explore these relationships. As team members move between projects, these graphs become more and more connected as team members collaborate and form relationships. We show that this connectivity, known as the ``small world effect," has a positive impact on team performance when the connectivity levels are moderate. Performance degrades, however, at both very high and very low levels of connectivity. This aligns with similar research findings of non-software teams.
\end{abstract}

@@

A ``small world'' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes\cite{watts1998collective}. ``Clustering'' in this context means how closely nodes in the graph are related to each other, or ``the friend of my friend is also my friend.'' ``Average path length'' is the the shortest route between all possible nodes, averaged over the entire graph. Milgram, in his seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}, or in other words, the average path length between any two people.

Uzzi \cite{uzzi2005collaboration} expanded on this by exploring how these small world networks correlate to collaboration and creativity. In his study, the network of artists involved in the making of Broadway musicals from 1945 to 1989 were studied. Performance was then measured based on financial success and critical acclaim, and statistically compared to the clustering and path length of the graph.

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} data sets of free and open source projects. Two datasets were analyzed -- data from from the popular open source project site \textit{Freecode}\footnote{http://freecode.com/}, and another from open source software repository \textit{SourceForge}\footnote{http://sourceforge.net/}. The Freecode dataset contained data up to September, 2013, while the SourceForge dataset was slightly older with data gathered on June 2009. Both datasets were chosen because they met two basic criteria: (1) they contained an exhaustive list of projects and contributors to those projects, and (2) they provided some sort of quantitative measure of project ``popularity." This popularity metric will be further described in sections below. Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the contributors and the edges representing one or more shared project between contributors. For our purposes, ``contributors'' includes any person that has been identified as being associated with the project by the project hosting site (Freecode and SourceForge, for our purposes). It is important to note that ``contributor'' does not always imply code contributions -- as issue reporting and documentation, among other things, are valid contributions. Various properties of the graph were analyzed, followed by a qualitative section to help link the quantitative analysis to subject matter expert experiences.

\subsubsection{Graph Structure}
For each dataset, we first considered the dataset to be one large graph. It was observed that the contributor graph was ``disconnected,'' meaning that there was not a path between all possible nodes. From this large disconnected graph, we can extract a set of connected subgraphs, where each connected subgraph was analyzed. Metrics were then computed for each of these subgraphs. In order to correctly compute the metrics below, a minimum subgraph size was required. Specifically, many of the graph metrics of interest measured ``triangles'' of nodes and their connections. This implies a subgraph of {$>= 3$} nodes is needed. Subgraphs of {$< 3$} nodes were not considered for analysis.

@@

A bipartite graph projection is necessary to further analyze the graph. A bipartite projection involves taking the disjoint node sets $P$ and $C$, and representing the graph as relationships between only one of those sets of nodes. A projection onto the contributor nodes, or $C$, of figure \ref{fig:example_bipartite_graph} is show in figure \ref{fig:example_bipartite_projection_graph}. Here, contributors are connected directly, with edges occurring if a two contributors share one or more common projects. It is important to note that the projection could have been done in terms of the projects, not the contributors. This would have led to a graph with nodes of projects $P$, each being linked by sharing a common contributor. This approach is a common way of representing these bipartite graphs\cite{newman2001scientific}, although it is known that bipartite projections are lossy compared to the data represented in the original graph\cite{zhou2007bipartite}.\\

@@

\subsubsection{Graph Analysis}
To support our hypothesis, there are several key metrics to be studied in regards to the aforementioned graph. Many of these metrics follow the work of Boccaletti \textit{et. al}\cite{boccaletti2006complex}, and other studies of similar graph types\cite{latora2001efficient,adamic1999small}.

@@

\noindent\\\textit{\textbf{Average Path Length ($L$)}}\\
Calculating the \textit{Path Length} ($L$) allowed us to determine how far removed contributors in the graph are from one another. The path length is calculated as the average distance between any two node pairs in the graph:

@@

Where {$v_i \in P$}, {$v_j \in P$}, and {$\lambda(v_i,v_j)$} denotes the shortest path between the two nodes. We only considered path length for connected subgraphs, avoiding some complexities of this calculation on disconnected graphs\cite{boccaletti2006complex}.

@@

\subsubsection{Qualitative Analysis}
So far we have explored, via data analysis, how closely project collaborators are connected to one another, and what correlations we can derive from this analysis. We have said little as to why these connections are made, how they are maintained, and what exactly a connection means at a personal level. Via qualitative analysis, we will attempt to enrich our data analysis by providing deeper context.\\
A connection between two people is much more than an edge between two nodes in a graph. It has been asserted that people tend to make connections with people with whom they share similarities\cite{mcpherson2001birds}. This idea of homophily, however, needs to be further expanded before we can make assertions about correlations to performance. It has been shown that demographic homophily is complex to measure, correlates little to performance outcomes, and in fact correlates negatively\cite{reagans2004make,lawrence1997perspective}. For our purposes, it is much more appropriate to examine how teams best leverage similarities of knowledge and information, as the flow of these assets can be a key driver to success\cite{nissen2002extended}.\\
In the same way we can explore the role of team familiarity. The two types of familiarity that we will explore are \textit{member} familiarity and \textit{task/project} familiarity\cite{harrison2003time}. These metrics, although not captured in data, could be important factors in \textit{how} teams form, and thus help us to understand \textit{why} teams perform the way that they do.\\

To further study these factors, we arranged interviews with two current professionals in the field. Both have had experience with project management and are certified Project Management Professionals (PMP)\textregistered. We proposed to them the following questions, with the intent of expanding further the questions posed above:

@@

Summary statistics for the graph analysis are captured in figure \ref{fig:summary_stats}.  Statistics analyzed included:\\
$Q$ - \textit{Small World Quotient}\\
$C^\Delta$ - \textit{Clustering Coefficient - Transitivity}\\
$\overline{C^\lambda}$ - \textit{Clustering Coefficient - Average}\\
$L$ - \textit{Average Path Length}\\
$N$ - \textit{Number of Nodes}\\
$E$ - \textit{Number of Edges}\\
The results shown were as measured over the number of projects $P$, with the total number of contributors $C$. The number of connected subgraphs analyzed is shown as $P_c$. The connected subgraphs are the main unit of analysis, and each statistical operation is applied to the set of subgraphs $P_c$.

A mean value of $Q$ was observed to be $> 1$ for both data sets, which meets the criteria for a small world network\cite{humphries2008network}. The median, however, is 1.0 in both cases, which does not match the small world network criteria. A possible explanation of this is the power law distribution of $Q$ observed in both data sets (figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}). This indicates that a large proportion of $Q$ values are around 1.\\

@@

Note that $P_c$ values are relatively small compared to $P$ and $C$ values. It would be intuitive to assume that {$N \times P_c \approx C$} -- in other words, if we split the entire graph into connected subgraphs, the number of subgraphs ($P_c$) multiplied by the average number of nodes ($N$) per subgraph should approximately equal the total number of contributors ($C$). This, however, is not the case. In fact, $P_c$ is notably smaller than expected. Recall that during analysis, subgraphs of {$< 3$} nodes were not considered. We can assume then that a large number of subgraphs are very small, with only one or two nodes. Further analysis is needed to fully understand the implications of this.

To validate the hypothesis, we expect to find that intermediate values of $Q$ will be correlated with more popular projects. To explore this possibility, we first compare the popularity given $Q$ values as shown in figures \ref{fig:q_fc_graph} and \ref{fig:q_sf_graph}. The $Q$ values do display the inverse U-shaped characteristic as compared to popularity, which aligns with the findings of Uzzi \textit{et. al}\cite{uzzi2005collaboration}.

@@

\subsection{Qualitative}
The research interviews provided a look into how some of these quantitative findings translate into actual industry practice. They also helped us to obtain a deeper understanding of what we had found in the data analysis, as we could begin to build some workplace context into our results. There were are few re-occurring themes from which we will base our analysis\cite{stmartin_interview,rana_interview}:

@@

Although our data analysis does not include any examination of a core/periphery, it is an interesting direction for future research. Notably, has been suggested that individual performance in a group is highest if that individual is not in the core or the periphery, but somewhere in between\cite{cattani2008core}. This is especially interesting as it aligns in principle with our hypothesis that intermediate values of $Q$ are optimal.

\noindent\\\textit{\textbf{Teams can stagnate without new members.}}\\
Both interview participants indicated that new members were essential in bringing new ideas to the team. A highly static team (or, a team with a high $Q$ value), runs the risk of falling behind the technology trends. Occasional new members (or, intermediate values of $Q$) were perceived as a positive in terms of information flow.

@@

\noindent\\\textit{\textbf{Priority, focus, and timeline are strong drivers.}}\\
Focus was specifically called out as an indicator of performance in one interview. Clear goals and a common objective were positive influencer, while continuously shifting focus was a negative influence. Research in this area suggests that creative teams have ``goals that are clear and compelling, but also open and challenging.''\cite{isaksen2002climate}

Timeline and priority were discussed as strong factors in team success. Team building could be thought of as function of the timeline -- meaning that as the timeline of the project increased, more time could be devoted to team building activities.

@@

\section{Future Research}
Social networks, interpersonal relationships, and software development are all complex topics. We have explored a narrow slice of these in context, and there are some important areas of future exploration. First, the contributor graph contains a large amount of subgraphs, and as a whole appears less clustered than some other observed social network graphs -- consistent with Madey's findings\cite{madey2002open} on SourceForge data. As noted in our analysis, a large number of these subgraphs have only one or two contributors. This implies that a large number of contributors work alone and on a single project. We do not, with the current analysis, know how (or if) this degrades the quality of our findings.\\
